Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays

### 2019-12-18 review_article.Rmd
formatting and image output

---
title: "Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
  word_document:
    toc: no
    toc_depth: '3'
link-citations: yes
fontsize: 11pt
csl: springer-vancouver.csl
yolo: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(cowplot)
library(png)
library(grid)
```

@@

## Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. 
Cancer maps and atlases dating back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

@@

<!--
Choropleth maps can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the ‘potentially more important results in the more populous communities’ that are geographically smaller [@SE]. 
Inset maps like in Brisbane city in Figure Figure 1<!--\ref{fig:choropleth-grid-create}-->c of the state of Queensland are commonly used to reduce distorted interpretations, but it is not always an appropriate solution. As this feature of population distributions continues to intensify, the need for cartograms as an alternative to a choropleth map should only increase.
-->

@@

## Tile Map
<!-- 173 words -->
A tile map provides a tessellated display of consistent shapes. 
A similar method to a rectangular cartogram, represents each geographic area using a square. The squares are tessellated to create a grid. This method has been used broadly in the media by FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR.
Each area is represented by a square of the same dimensions, each tile is usually one unit of measurement, this could be geographic regions such as states or population-based that use a consistent measure of population for each tile. Regions with over four neighbors require some necessary displacement. The tile map uses color to represent a value of a statistic for each area. A similar method to a rectangular cartogram represents each geographic area using a square of the same dimensions. The squares tessellate to create a grid. There are online media sources using this method, these include [NPR](https://www.npr.org/2016/10/18/498406765/npr-battleground-map-hillary-clinton-is-winning-and-its-not-close), [FiveThirtyEight](https://projects.fivethirtyeight.com/2016-election-forecast/), [The Wall Street Journal](http://graphics.wsj.com/elections/2016/2016-electoral-college-map-predictions/), [The Washington Post](https://www.washingtonpost.com/graphics/politics/2016-election/50-state-poll/). Tile maps may be difficult to create as they are best created manually, they require additional time and care as the number of geographic areas to include increases.
<!-- cite mosaic cartograms -->

@@

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features, found in online cancer maps, include tool tips, drop-down menus, data selection, zooming, and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports can be found in various online cancer maps [@roberts2019communication].

@@

Figure 7 <!--\ref{fig:animating}--> shows two examples of more sophisticated interactive maps. The Spanish Cancer map (left) contains a linked display between a choropleth map and time series plots of cancer change. In linked plots, changing values in one display will trigger changes of corresponding elements in another display. Here, the temporal change in the choropleth map can be played out as an animation. Mousing over the time series plots will highlight the line for a particular region. The Canadian Breast Cancer Mortality map (right) has a magnifying glass that allows the user to zoom into small areas. It is easy to control and shows precise details in small areas.

>>>>

---
title: "Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
    dev: jpeg
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{fancyhdr}\pagestyle{fancy}\fancyhead[R]{\thepage}
link-citations: yes
fontsize: 11pt
geometry: margin = 1.2in
csl: springer-vancouver.csl
yolo: yes
---

```{r setup, include=FALSE}
fig_path <- glue::glue("figs/")
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.path = fig_path)
library(ggplot2)
library(cowplot)
library(png)
library(grid)
```

@@

## Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization.Cancer maps and atlases dating back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

@@

<!--
Choropleth maps can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the ‘potentially more important results in the more populous communities’ that are geographically smaller [@SE]. 
Inset maps like in Brisbane city in Figure Figure 1<!--\ref{fig:choropleth-grid-create}c of the state of Queensland are commonly used to reduce distorted interpretations, but it is not always an appropriate solution. As this feature of population distributions continues to intensify, the need for cartograms as an alternative to a choropleth map should only increase.
-->

@@

## Tile Map
<!-- 173 words -->
A tile map provides a tessellated display of consistent shapes.  A similar method to a rectangular cartogram, represents each geographic area using a square. The squares are tessellated to create a grid. Each area is represented by a square of the same dimensions, each tile is usually one unit of measurement, this could be geographic regions such as states or population-based that use a consistent measure of population for each tile. Regions with over four neighbors require some necessary displacement. The tile map uses color to represent a value of a statistic for each area. A similar method to a rectangular cartogram represents each geographic area using a square of the same dimensions. There are online media sources using this method, these include [NPR](https://www.npr.org/2016/10/18/498406765/npr-battleground-map-hillary-clinton-is-winning-and-its-not-close), [FiveThirtyEight](https://projects.fivethirtyeight.com/2016-election-forecast/), [The Wall Street Journal](http://graphics.wsj.com/elections/2016/2016-electoral-college-map-predictions/), [The Washington Post](https://www.washingtonpost.com/graphics/politics/2016-election/50-state-poll/). Tile maps may be difficult to create as they are best created manually, they require additional time and care as the number of geographic areas to include increases.
<!-- cite mosaic cartograms -->

@@

Interactivity enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features, found in online cancer maps, include tool tips, drop-down menus, data selection, zooming, and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports can be found in various online cancer maps and are shown in Figure 5 [@roberts2019communication].

@@

Figure 6 <!--\ref{fig:animating}--> shows two examples of more sophisticated interactive maps. The Spanish Cancer map (left) contains a linked display between a choropleth map and time series plots of cancer change. In linked plots, changing values in one display will trigger changes of corresponding elements in another display. Here, the temporal change in the choropleth map can be played out as an animation. Mousing over the time series plots will highlight the line for a particular region. The Canadian Breast Cancer Mortality map (right) has a magnifying glass that allows the user to zoom into small areas. It is easy to control and shows precise details in small areas.
